On the propriety of restricted Boltzmann machines

### 2016-11-17 writing/draft.Rmd
formatting

```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size. Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=4}
three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.degen

@@

```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
p.elpr
```

```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$. Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=4}
p.exp_diff
```

@@

```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$. For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases also.", fig.height=5}
plot_dat_grid %>% 
  mutate(dist = sqrt(r1^2 + r2^2)/(H + V + H*V)) %>%
  group_by(dist, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n()) %>%
  mutate(Hiddens = as.character(H), Visibles = paste0("paste(n[v], '= ", V, "')")) %>%
  ggplot() +
  geom_point(aes(dist, frac_degen, colour = Hiddens), size = .5) +
  #geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
  facet_wrap(~Visibles, labeller = label_parsed) +
  ylim(c(0,1)) +
  xlab(expression(paste("||", theta, "||", "/(", n[h], " + ", n[v], " + ", n[h], "*", n[v], "), Average distance from the origin"))) +
  ylab("Fraction Near-degenerate") +
  scale_colour_discrete(expression(n[h]))
```

@@

sample.params %>% 
  ungroup %>% 
  filter(!near_hull) %>% select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
  rename(`$\\theta_{v1}$` = v1,
         `$\\theta_{v2}$` = v2,
         `$\\theta_{v3}$` = v3,
         `$\\theta_{v4}$` = v4,
         `$\\theta_{h1}$` = h1,
         `$\\theta_{h2}$` = h2,
         `$\\theta_{h3}$` = h3,
         `$\\theta_{h4}$` = h4,
         `$\\theta_{11}$` = theta11,
         `$\\theta_{12}$` = theta12,
         `$\\theta_{13}$` = theta13,
         `$\\theta_{14}$` = theta14,
         `$\\theta_{21}$` = theta21,
         `$\\theta_{22}$` = theta22,
         `$\\theta_{23}$` = theta23,
         `$\\theta_{24}$` = theta24,
         `$\\theta_{31}$` = theta31,
         `$\\theta_{32}$` = theta32,
         `$\\theta_{33}$` = theta33,
         `$\\theta_{34}$` = theta34,
         `$\\theta_{41}$` = theta41,
         `$\\theta_{42}$` = theta42,
         `$\\theta_{43}$` = theta43,
         `$\\theta_{44}$` = theta44) %>%
  gather("Parameter", "Value") %>%
  mutate(Value = paste0("$", Value, "$")) -> tbl

cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$. This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", digits = 10)
```

@@

marginal_means %>%
  group_by(v1, v2, v3, v4, image_id) %>%
  do(data.frame(C = 2000^(1/3)*var(.$means[[1]]))) %>%
  left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
  mutate(M_eff = sigma2/C,
         model = "Marginal Likelihood") %>%
  bind_rows(trunc_means %>%
    group_by(v1, v2, v3, v4, image_id) %>%
    do(data.frame(C = 5000^(1/3)*var(.$means[[1]]))) %>%
    left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
    mutate(M_eff = sigma2/C,
           model = "Truncated Normal Prior")) %>%
  select(-C, -sigma2) %>%
  spread(model, M_eff) %>%
  ungroup() %>%
  select(-(v1:v4)) %>%
  gather(Model, `$M_{eff}$`, -image_id) %>%
  mutate(`$M_{eff}$` = 1000*`$M_{eff}$`) %>%
  spread(image_id, `$M_{eff}$`) -> M_eff_table

@@

```{r m_eff_table}
M_eff_table %>%
  gather(Image, M_eff, -Model) %>%
  spread(Model, M_eff) %>%
  mutate(Vector = as.numeric(Image)) %>%
  arrange(Vector) -> M_eff_tmp

@@

The effective sample sizes for a chain of length $M = 1000$ of all $2^4 = 16$ possible vectors of visibles are presented in Table \ref{tab:m_eff}. These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws. Thus, BwTNLV would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution. For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BeTNLV does not seem to be an effective method for fitting the RBM of computing resources are at all limited.

@@

all_statistics %>%
  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
  filter(grepl("image", statistic) & !is.na(value)) %>%
  separate(statistic, into = c("junk", "statistic")) %>%
  mutate(statistic = factor(statistic, labels = paste("Vector", 1:length(unique(statistic))))) %>%
  left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
  ggplot() +
  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
  geom_vline(aes(xintercept = prob)) +
  geom_vline(aes(xintercept = prob_emp), colour = "red") +
  facet_wrap(~statistic, scales = "free_x") + 
  ylab("Scaled Posterior Density") + xlab("Probability of Vector of Visibles") +
  scale_colour_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
  scale_fill_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
  theme(legend.position = "bottom") -> p.models
```

```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML. The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to simulate the training data while the red vertical lines show the empirical distribution. BwTNML produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
p.models
```

>>>>

```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size. Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=3.5, fig.align="ht!"}
three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.degen

@@

```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size. As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
p.elpr
```

```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$. Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=3.5, fig.align="ht!"}
p.exp_diff
```

@@

```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$. For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases also.", fig.height=4, fig.align="ht!"}
plot_dat_grid %>% 
  mutate(dist = sqrt(r1^2 + r2^2)/(H + V + H*V)) %>%
  group_by(dist, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n()) %>%
  mutate(Hiddens = as.character(H), Visibles = paste0("paste(n[v], '= ", V, "')")) %>%
  ggplot() +
  geom_point(aes(dist, frac_degen, colour = Hiddens), size = .5) +
  #geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
  facet_wrap(~Visibles, labeller = label_parsed) +
  ylim(c(0,1)) +
  xlab(expression(paste("||", theta, "||", "/(", n[h], " + ", n[v], " + ", n[h], "*", n[v], "), Average distance from the origin"))) +
  ylab("Fraction Near-degenerate") +
  scale_colour_discrete(expression(n[h]))
```

@@

old_pen <- options()$scipen
options("scipen" = 100)
sample.params %>% 
  ungroup %>% 
  filter(!near_hull) %>% select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
  rename(`$\\theta_{v1}$` = v1,
         `$\\theta_{v2}$` = v2,
         `$\\theta_{v3}$` = v3,
         `$\\theta_{v4}$` = v4,
         `$\\theta_{h1}$` = h1,
         `$\\theta_{h2}$` = h2,
         `$\\theta_{h3}$` = h3,
         `$\\theta_{h4}$` = h4,
         `$\\theta_{11}$` = theta11,
         `$\\theta_{12}$` = theta12,
         `$\\theta_{13}$` = theta13,
         `$\\theta_{14}$` = theta14,
         `$\\theta_{21}$` = theta21,
         `$\\theta_{22}$` = theta22,
         `$\\theta_{23}$` = theta23,
         `$\\theta_{24}$` = theta24,
         `$\\theta_{31}$` = theta31,
         `$\\theta_{32}$` = theta32,
         `$\\theta_{33}$` = theta33,
         `$\\theta_{34}$` = theta34,
         `$\\theta_{41}$` = theta41,
         `$\\theta_{42}$` = theta42,
         `$\\theta_{43}$` = theta43,
         `$\\theta_{44}$` = theta44) %>%
  gather("Parameter", "Value") %>%
  mutate(Value = paste0("$", round(Value, 10), "$")) -> tbl
options("scipen" = old_pen)

cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$. This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = rep(c("l", "r"), times = 3))
```

@@

marginal_means %>%
  group_by(v1, v2, v3, v4, image_id) %>%
  do(data.frame(C = 2000^(1/3)*var(.$means[[1]]))) %>%
  left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
  mutate(M_eff = sigma2/C,
         model = "BwTNML") %>%
  bind_rows(trunc_means %>%
    group_by(v1, v2, v3, v4, image_id) %>%
    do(data.frame(C = 5000^(1/3)*var(.$means[[1]]))) %>%
    left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
    mutate(M_eff = sigma2/C,
           model = "BwTNLV")) %>%
  select(-C, -sigma2) %>%
  spread(model, M_eff) %>%
  ungroup() %>%
  select(-(v1:v4)) %>%
  gather(Model, `$M_{eff}$`, -image_id) %>%
  mutate(`$M_{eff}$` = 1000*`$M_{eff}$`) %>%
  spread(image_id, `$M_{eff}$`) -> M_eff_table

@@

```{r m_eff_table}
M_eff_table %>%
  gather(Image, M_eff, -Model) %>%
  spread(Model, M_eff) %>%
  mutate(Vector = as.numeric(Image)) %>%
  select(Vector, BwTNML, BwTNLV) %>%
  arrange(Vector) -> M_eff_tmp

@@

The effective sample sizes for a chain of length $M = 1000$ of all $2^4 = 16$ possible vectors of visibles are presented in Table \ref{tab:m_eff}. These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws. Thus, BwTNLV would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution. For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM of computing resources are at all limited.

@@

all_statistics %>%
  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
  filter(grepl("image", statistic) & !is.na(value)) %>%
  separate(statistic, into = c("junk", "statistic")) %>%
  mutate(statistic = factor(statistic, labels = paste("Vector", 1:length(unique(statistic))))) %>%
  left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
  ggplot() +
  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
  geom_vline(aes(xintercept = prob)) +
  geom_vline(aes(xintercept = prob_emp), colour = "red") +
  facet_wrap(~statistic, scales = "free_x") + 
  ylab("Scaled Posterior Density") + xlab("Probability of Vector of Visibles") +
  scale_colour_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
  scale_fill_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
  theme(legend.position = "bottom") -> p.models

sx <- scale_x_continuous()
sx$trans$breaks <- function(range) pretty(range, n = 3)
```

```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML. The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to simulate the training data while the red vertical lines show the empirical distribution. BwTNML produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
p.models + sx
```
